---
title: "singIST Vignette"
author: "Aitor Moruno-Cuenca"
output: BiocStyle::html_document
vignette: > 
  %\VignetteIndexEntry{singIST Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

This Vignette demonstrates how to use the singIST package. singIST is a 
computational method for comparative single-cell transcriptomics between
organisms and experimental conditions. Its cornerstone use case is comparing
scRNA-seq profiles between disease models and human diseases. Throughout this 
vignette we will work with the following data:

\begin{itemize}
\item "cytokine_counts.rda" Contains Human scRNAseq pseudobulk lognormalized
count matrix. The pseudobulk is aggregated by sample and cell type and contains
all genes belonging to the pathway Cytokine-Cytokine receptor interaction [KEGG]
\item "DC_counts.rda" Contains Human scRNAseq pseudobulk lognormalized
count matrix. The pseudobulk is aggregated by sample and cell type and contains
all genes belonging to the pathway Dendritic Cells in regulating Th1/Th2
Development [BIOCARTA]
\item "OXA.rda" 
\end{itemize}

```{r}
library(singIST)
# Example code
utils::sessionInfo()
```

# Initialize objects for analysis
## Superpathway input 

### Pathway: Dendritic Cells in Th1/Th2 Development [BIOCARTA]

```{r}
# Initialize pathway object
dc_pathway <- methods::new("pathway",
                            standard_name = "BIOCARTA_DC_PATHWAY",
                            dbsource = "BIOCARTA",
                            collection = "c2",
                            subcollection = "CP")

# We now define the superpathway gene sets object in which we define the 
# cell types we are interested in modelling
dc_superpathway <- methods::new("superpathway.gene.sets",
                                pathway_info = dc_pathway,
                                celltypes = c("Dendritic Cells", 
                                                "Langerhan Cells", 
                                                "T-cell"))

# Load a Gene Set Collection (gse) object from MsigDB for homo sapiens and 
# HGNC gene symbols 
gse <- msigdb::getMsigdb(org = "hs", id = "SYM",
                            version = msigdb::getMsigdbVersions()[1])
# We choose the same gene sets for all the cell types, hence we use the
# method setRepeatGeneSets that pulls the gene set associated to the pathway
# and sets the same gene set for all cell types
dc_superpathway <- setRepeatGeneSets(dc_superpathway, gse = gse) 
print(slot(dc_superpathway, "gene_sets_celltype"))
```

### Pathway: Cytokine-Cytokine receptor interaction [KEGG]

```{r}
# Initialize pathway object
cytokine_pathway <- methods::new(
    "pathway", standard_name = "KEGG_CYTOKINE_CYTOKINE_RECEPTOR_INTERACTION",
    dbsource = "KEGG", collection = "c2", subcollection = "CP")
# Superpathway gene sets object
cytokine_superpathway <- methods::new(
    "superpathway.gene.sets", pathway_info = cytokine_pathway,
    celltypes = c("Dendritic Cells", "Keratinocytes", "Langerhan Cells",
                    "Melanocytes", "T-cell"))
# Pull Gene Set 
gene_set <- pullGeneSet(cytokine_pathway, gse = gse)
# Set different gene sets per cell type randomly. This would exemplify an
# ad-hoc pathway where we set different gene sets per cell type 
slot(cytokine_superpathway, "gene_sets_celltype") <- list(
    sample(gene_set, 50), sample(gene_set, 100), sample(gene_set, 10),
    sample(gene_set, 60), sample(gene_set, 160))
```

## Human scRNAseq input 

### Pathway: Dendritic Cells in Th1/Th2 Development [BIOCARTA]

```{r}
# Load the pseudobulk lognormalized count matrix for the Human Dendritic Cells
# Th1/Th2 development [BIOCARTA] genes
counts_dc <- DC_counts
# Remove the cell types as we are not going to model
counts_dc <- counts_dc[!(sub("_.*$", "", rownames(counts_dc)) %in% 
                        c("MastC-other", "Keratinocytes", "Melanocytes")), ]

# Initialize hyperparameter object 
# Choose the array of quantile combinations that will be tested in the Cross
# Validation. Ideally one should choose a wide range of quantiles to test,
# as a narrow selection might not find the globally optimal model. The more
# quantiles one chooses the more computation time it will take, for the sake
# of simplicity we choose a small amount of quantiles to test
quantile_comb_table <- as.matrix(
    RcppAlgos::permuteGeneral(seq(0.1, 0.95, by = 0.3),
                                m = length(slot(dc_superpathway, "celltypes")), 
                                TRUE))

# We choose a "binary" outcome, we will test at maximum 3 PLS components,
# and we choose a LOOCV (1 fold), since we choose LOOCV the number of
# repetitions arguments will also be 1 by default
dc_hyperparameters <- methods::new("hyperparameters",
                                    quantile_comb_table = quantile_comb_table,
                                    outcome_type = "binary",
                                    number_PLS = as.integer(3),
                                    folds_CV = as.integer(1),
                                    repetition_CV = as.integer(1))

# Initialize the superpathway input object that will be used to fit the
# optimal asmbPLSDA
# Initialize superpathway input object
# Sample_id contains the identifiers of each sample as shown in
# the rownames of the pseudobulk lognormalized matrix, the sample class
# indicates the class each sample belongs to
dc_superpathway_input <- methods::new(
    "superpathway.input", superpathway_info = dc_superpathway,
    hyperparameters_info = dc_hyperparameters, pseudobulk_lognorm = counts_dc,
    sample_id = c("AD1", "AD13", "AD2", "AD3","AD4", "HC1-2", "HC3", "HC4",
                    "HC5"),
    sample_class = c(rep("AD",5), rep("HC", 4)),
    base_class = "HC",
    target_class = "AD")
```

### Pathway: Cytokine-Cytokine receptor interaction [KEGG]

```{r}
# Load the pseudobulk lognormalized count matrix
counts_cytokine <- cytokine_counts
# Remove the cell types as we are not going to model
counts_cytokine <- counts_cytokine[!(sub("_.*$", "",
                                    rownames(counts_cytokine)) %in% 
                                    c("MastC-other")), ]

# Initialize hyperparameter object 
quantile_comb_table <- as.matrix(
    RcppAlgos::permuteGeneral(seq(0.05, 0.95, by = 0.2),
                                m = length(slot(cytokine_superpathway, 
                                                "celltypes")), 
                                TRUE))

# We choose a "binary" outcome, we will test at maximum 2 PLS components,
# and we choose a 5-Fold CV with 5 repetitions each
cytokine_hyperparameters <- methods::new("hyperparameters",
                                    quantile_comb_table = quantile_comb_table,
                                    outcome_type = "binary",
                                    number_PLS = as.integer(2),
                                    folds_CV = as.integer(5),
                                    repetition_CV = as.integer(5))

# Initialize the superpathway input object that will be used to fit the
# optimal asmbPLSDA
cytokine_superpathway_input <- methods::new(
    "superpathway.input", superpathway_info = cytokine_superpathway,
    hyperparameters_info = cytokine_hyperparameters,
    pseudobulk_lognorm = counts_cytokine,
    sample_id = c("AD1", "AD13", "AD2", "AD3","AD4", "HC1-2", "HC3", "HC4",
                    "HC5"),
    sample_class = c(rep("AD",5), rep("HC", 4)),
    base_class = "HC",
    target_class = "AD")
```

## Oxazolone scRNAseq input

```{r}
# Load the disease model Seurat object scRNA-seq
oxazolone <- OXA
# Rename the variable containing the class each sample belongs to and the
# cell type cluster variable into "class" and "celltype_cluster" names as
# requested by mapping.organism object 
colnames(slot(oxazolone, "meta.data"))[c(5,11)] <- c("class", 
                                                        "celltype_cluster")
# Initialize the mapping.organism object that contains the information on
# the disease model
# The target_class is the element of variable "class" that serves as the
# target class and likewise for base_class. The mapping.organism object
# should contain the cell types used in all human superpathways under study
oxa_org <- new("mapping.organism", organism = "Mus musculus", 
                target_class = "OXA", base_class = "ETOH",
                celltype_mapping = list(
                    "Dendritic Cells" = c("cDC2", "cDC1", "migratory DCs"),
                    "Keratinocytes" = c("Keratinocytes"),
                    "Langerhan Cells" = c("LC"), 
                    "Melanocytes" = c(),
                    "T-cell" = c("DETC", "dγdT", "T")),
                counts = oxazolone
)
```

# Fit Optimal asmbPLS-DA and compute validation metrics for the optimal model

```{r}
# One can choose to either fit each superpathway individually with fitOptimal or
# use its wrapper multiple_fitOptimal to fit them all at once. 
# enable parallellization with 2 workers
library(BiocParallel)
# SnowParam works with Windows, if you work with Linux MulticoreParam is better
BiocParallel::register(BiocParallel::SnowParam(workers = 2,
                        exportglobals = FALSE, progressbar = TRUE),
                        default = TRUE)  
superpathways <- list(dc_superpathway_input, cytokine_superpathway_input)
superpathways_fit <- multiple_fitOptimal(
    superpathways, parallel = c(TRUE, FALSE), npermut = c(100),
    type = c("jackknife","subsampling"), nsubsampling = c(NULL, 5))
# disable parallelization
BiocParallel::register(BiocParallel::SerialParam(), default = TRUE) 
```
# Compute recapitulation metrics

```{r}
# singISTrecapitulations computes the recapitulations for the superpathway
# between the fitted optimal model and the disease model. It automatically
# computes the biological link function, the singIST treated samples, and
# derives the scores to compute the recapitulations. By default it assumes
# the organism for which the asmbPLSDA model was trained is "hsapiens", it 
# also assumes that the gene annotation for the mapped organism is 
# "external_gene_name". These parameters are by default but other cases
# can be considered by specifying the parameters "from_to" and 
# "object_gene_identifiers"
recapitulations_dc <- singISTrecapitulations(oxa_org, superpathways_fit[[1]])
recapitulations_cytokine <- singISTrecapitulations(oxa_org,
                                                    superpathways_fit[[2]])
```

```{r}
# One can also compute multiple singIST recapitulations for the same disease
# model across multiple superpathways of interest by using 
# multiple_singISTrecapitulations
recapitulations_multiple <- multiple_singISTrecapitulations(oxa_org,
                                                            superpathways_fit)
```


